---
title: "homework4"
author: "Xueyuan Li"
date: "2023-02-22"
output: github_document
---

```{r}
library('MASS') ## for 'mcycle'
library('manipulate') ## for 'manipulate'
library('caret') ## 'knnreg' and 'createFolds'
library('ggplot2')
```

Goal: Understand and implement various ways to approximate test error.
In the ISLR book, read section 6.1.3 “Choosing the Optimal Model” and section 5.1 “Cross-Validation”. Extend and convert the attached effective-df-aic-bic-mcycle.R R script into an R markdown file that accomplishes the following tasks.

### 1,Randomly split the mcycle data into training (75%) and validation (25%) subsets.

```{r}
require(caTools)
set.seed(101) 
sample_y = sample.split(mcycle$accel, SplitRatio = .75)
train_y = subset(mcycle$accel, sample_y == TRUE)
val_y  = subset(mcycle$accel, sample_y == FALSE)

```

```{r}
sample_x = sample.split(mcycle$times, SplitRatio = .75)
train_x = subset(mcycle$times, sample_x == TRUE)
val_x  = subset(mcycle$times, sample_x == FALSE)

train_x=matrix(train_x) 
train_y=matrix(train_y)
val_x=matrix(val_x)
val_y=matrix(val_y)

```


### 2, Using the mcycle data, consider predicting the mean acceleration as a function of time. Use the Nadaraya-Watson method with the k-NN kernel function to create a series of prediction models by varying the tuning parameter over a sequence of values. (hint: the script already implements this)



```{r}

## Epanechnikov kernel function
## x  - n x p matrix of training inputs
## x0 - 1 x p input where to make prediction
## lambda - bandwidth (neighborhood size)
kernel_epanechnikov <- function(x, x0, lambda=1) {
  d <- function(t)
    ifelse(t <= 1, 3/4*(1-t^2), 0)
  z <- t(t(x) - x0)
  d(sqrt(rowSums(z*z))/lambda)
}

## k-NN kernel function
## x  - n x p matrix of training inputs
## x0 - 1 x p input where to make prediction
## k  - number of nearest neighbors
kernel_k_nearest_neighbors <- function(x, x0, knn=1) {
  ## compute distance betwen each x and x0
  z <- t(t(x) - x0)
  d <- sqrt(rowSums(z*z))

  ## initialize kernel weights to zero
  w <- rep(0, length(d))
  
  ## set weight to 1 for k nearest neighbors
  w[order(d)[1:knn]] <- 1
  
  return(w)
}

## Make predictions using the NW method
## y  - n x 1 vector of training outputs
## x  - n x p matrix of training inputs
## x0 - m x p matrix where to make predictions
## kern  - kernel function to use
## ... - arguments to pass to kernel function
nadaraya_watson <- function(y, x, x0, kern, ...) {
  k <- t(apply(x0, 1, function(x0_) {
    k_<- kern(x, x0_, ...)
    k_/sum(k_)
  }))
  yhat <- drop(k %*% y)
  attr(yhat, 'k') <- k
  return(yhat)
}

## Helper function to view kernel (smoother) matrix
matrix_image <- function(x) {
  rot <- function(x) t(apply(x, 2, rev))
  cls <- rev(gray.colors(20, end=1))
  image(rot(x), col=cls, axes=FALSE)
  xlb <- pretty(1:ncol(x))
  xat <- (xlb-0.5)/ncol(x)
  ylb <- pretty(1:nrow(x))
  yat <- (ylb-0.5)/nrow(x)
  axis(3, at=xat, labels=xlb)
  axis(2, at=yat, labels=ylb)
  mtext('Rows', 2, 3)
  mtext('Columns', 3, 3)
}

## Compute effective df using NW method
## y  - n x 1 vector of training outputs
## x  - n x p matrix of training inputs
## kern  - kernel function to use
## ... - arguments to pass to kernel function
effective_df <- function(y, x, kern, ...) {
  y_hat <- nadaraya_watson(y, x, x,
    kern=kern, ...)
  sum(diag(attr(y_hat, 'k')))
}

## loss function
## y    - train/test y
## yhat - predictions at train/test x
loss_squared_error <- function(y, yhat)
  (y - yhat)^2

## test/train error
## y    - train/test y
## yhat - predictions at train/test x
## loss - loss function
error <- function(y, yhat, loss=loss_squared_error)
  mean(loss(y, yhat))

## AIC
## y    - training y
## yhat - predictions at training x
## d    - effective degrees of freedom
aic <- function(y, yhat, d)
  error(y, yhat) + 2/length(y)*d

## BIC
## y    - training y
## yhat - predictions at training x
## d    - effective degrees of freedom
bic <- function(y, yhat, d)
  error(y, yhat) + log(length(y))/length(y)*d


## make predictions using NW method at training inputs
y_hat <- nadaraya_watson(train_y, train_x, train_x,
  kernel_epanechnikov, lambda=5)

## view kernel (smoother) matrix
matrix_image(attr(y_hat, 'k'))

## compute effective degrees of freedom
edf <- effective_df(train_y, train_x, kernel_epanechnikov, lambda=5)
edf 
```

```{r}


plot(train_x,train_y, xlab="Time (ms)", ylab="Acceleration (g)")
lines(train_x, y_hat, col="#882255", lwd=2) 


AIC<-aic(train_y,y_hat,edf)
BIC<-bic(train_y,y_hat,edf)

AIC
BIC

```






### 3, With the squared-error loss function, compute and plot the training error, AIC, BIC, and validation error (using the validation data) as functions of the tuning parameter.
```{r}

# Create a sequence of tuning parameters
k <- seq(1, 20, by = 1)
length(k)
train_error<-matrix(0, nrow = length(k), ncol = 1)
val_error<- matrix(0, nrow = length(k), ncol = 1)
AIC <- matrix(0, nrow = length(k), ncol = 1)
BIC<- matrix(0, nrow = length(k), ncol = 1)
train=data.frame(train_x,train_y)
val=data.frame(val_x, val_y)


for (i in 1:20) {
  i
  y_hat_train <-nadaraya_watson(train_y, train_x, train_x, kernel_k_nearest_neighbors,knn=i)
  y_hat_val <- nadaraya_watson(val_y, val_x, val_x, kernel_k_nearest_neighbors, knn=i)

  # Calculate the training error
  train_error[i] <- error(train_y, y_hat_train ,loss=loss_squared_error)
  
  # Calculate the validation error
  val_error[i] <- error(val_y, y_hat_val, loss=loss_squared_error)
  
  
  #d <- effective_df(train_y,train_x, tuning_seq[i])
  #train_edf <- effective_df(train_y,train_x, kernel_epanechnikov, lambda=5)
  AIC[i] <- aic(train_y, y_hat_train , edf)
  BIC[i] <- bic(train_y, y_hat_train , edf)

}


train_error
val_error
AIC
BIC

# Plot the errors as a function of the tuning parameter
par(mfrow = c(2, 2))
plot(k, train_error, type = "l", xlab = "Tuning Parameter", ylab = "Training Error")
#abline(v = tuning_seq[which.min(train_error)], lty = 2)
plot(k, AIC, type = "l", xlab = "Tuning Parameter", ylab = "AIC")
#abline(v = tuning_seq[which.min(AIC)], lty = 2)
plot(k, BIC, type = "l", xlab = "Tuning Parameter", ylab = "BIC")
#abline(v = tuning_seq[which.min(BIC)], lty = 2)
plot(k,val_error, type = "l", xlab = "Tuning Parameter", ylab = "Validation Error")
#abline(v = tuning_seq[which.min(val_error)], lty = 2)


```




### 4, For each value of the tuning parameter, Perform 5-fold cross-validation using the combined training and validation data. This results in 5 estimates of test error per tuning parameter value.
```{r}
## 5-fold cross-validation of knnreg model
## create five folds
set.seed(1985)
inc_flds  <- createFolds(mcycle$accel, k=5)
print(inc_flds)
sapply(inc_flds, length)  ## not all the same length

cvknnreg <- function(kNN = 10, flds=inc_flds) {
  cverr <- rep(NA, length(flds))
  for(tst_idx in 1:length(flds)) { ## for each fold
    
    ## get training and testing data
    #train=data.frame(train_x,train_y)
    #val=data.frame(val_x, val_y)
    #inc_trn <- train
    #inc_tst <- val
    
    ## get training and testing data
    inc_trn <- mcycle[-flds[[tst_idx]],]
    inc_tst <- mcycle[ flds[[tst_idx]],]
    
    ## fit kNN model to training data
    knn_fit <- knnreg(accel ~ times, k=kNN, data=inc_trn)
    
    ## compute test error on testing data
    pre_tst <- predict(knn_fit, inc_tst)
    
    cverr[tst_idx] <- mean((inc_tst$accel - pre_tst)^2)
    
  }
  
  return(cverr)
}

cverrs <- sapply(1:20, cvknnreg)
print(cverrs) 
cverrs_mean <- apply(cverrs, 2, mean)
cverrs_sd   <- apply(cverrs, 2, sd)

cverrs_mean
cverrs_sd 

## Plot the results of 5-fold CV for kNN = 1:20
plot(x=1:20, y=cverrs_mean, 
     ylim=range(cverrs),
     xlab="'k' in kNN", ylab="CV Estimate of Test Error")
segments(x0=1:20, x1=1:20,
         y0=cverrs_mean-cverrs_sd,
         y1=cverrs_mean+cverrs_sd,ylim=range(cverrs))
best_idx <- which.min(cverrs_mean)
points(x=best_idx, y=cverrs_mean[best_idx], pch=20)
abline(h=cverrs_mean[best_idx] + cverrs_sd[best_idx], lty=3)


```



### 5, Plot the CV-estimated test error (average of the five estimates from each fold) as a function of the tuning parameter. Add vertical line segments to the figure (using the segments function in R) that represent one “standard error” of the CV-estimated test error (standard deviation of the five estimates from each fold).
```{r}
## Plot the results of 5-fold CV for kNN = 1:20
plot(x=1:20, y=cverrs_mean, 
     ylim=range(cverrs),
     xlab="'k' in kNN", ylab="CV Estimate of Test Error")
segments(x0=1:20, x1=1:20,
         y0=cverrs_mean-cverrs_sd,
         y1=cverrs_mean+cverrs_sd,ylim=range(cverrs))
best_idx <- which.min(cverrs_mean)
points(x=best_idx, y=cverrs_mean[best_idx], pch=20)
abline(h=cverrs_mean[best_idx] + cverrs_sd[best_idx], lty=3)
```


### 6, Interpret the resulting figures and select a suitable value for the tuning parameter.

The resulting figure shows the CV-estimated test error as a function of the tuning parameter for the k-nearest neighbor regression model. The y-axis represents the mean CV-estimated test error, and the x-axis represents the values of the tuning parameter (k).

The figure also includes vertical line segments that represent one standard error of the CV-estimated test error. These segments give an indication of the variability of the CV-estimated test error across the different folds.

To select a suitable value for the tuning parameter, we look for the value of k=14 that has the lowest CV-estimated test error. In the figure, this corresponds to the point with the lowest y-value. Additionally, we want to choose a value of k that is not too high, as this could result in overfitting, nor too low, as this could result in underfitting.


To select a suitable value for the tuning parameter, we look for the value of k that has the lowest CV-estimated test error. In the figure, this corresponds to the point with the lowest y-value. Additionally, we want to choose a value of k that is not too high, as this could result in overfitting, nor too low, as this could result in underfitting.

